---
title: "Working_GSEA_RUN_Pre_Cyst"
author: "Ryan Brownlee"
date: '2022-06-23'
output: html_document
---
Load Libraries 
```{r}
library(TFEA.ChIP)
library(DESeq2)
library(biomaRt)
library(dplyr)
library(plotly)
```

Load in vignette data
#This is just if you want to test the vignette if getting used to TFEA.ChIP

```{r}
data( "hypoxia_DESeq", "hypoxia", package="TFEA.ChIP" ) # load example datasets
hypoxia_table <- preprocessInputData( hypoxia_DESeq )
head(hypoxia_table)
```
Load in our Pre-Cyst data
```{r}
alldegs <- read.csv("/Users/mattr/Documents/Lasseigne_Lab_KURE/deseq2_fullres_39.csv")
```
Remove NAs from alldegs object (Our Data)
```{r}
alldegs <- tidyr::drop_na(alldegs, log2FoldChange) #GSEA_RUN won't work w/o this step.
```

```{r}
chip_index <- get_chip_index( encodeFilter = TRUE ) #GSEA_RUN will compare our dataset to the encode database of transcription factors
```
Mouse to human (ensembl)
```{r}
ens_annot <- read.csv("/Users/mattr/Documents/Lasseigne_Lab_KURE/annot_ens_humanmouse.csv")
  #conversion table
degs_hom <- merge(alldegs, ens_annot, by.x = "X", by.y = "Mouse.gene.stable.ID")
  #biomart to convert human ens to human entrez
human <- useEnsembl("ensembl", dataset = "hsapiens_gene_ensembl", mirror = "useast")
#human ensembl to human entrez mapping
genes <- getBM(filters = "ensembl_gene_id", attributes = c("ensembl_gene_id","entrezgene_id"), values = degs_hom$Gene.stable.ID, mart = human)
```
Merge entrez ID's back with differential expression data
```{r}
entrezdegs <- merge(degs_hom, genes, by.x = "Gene.stable.ID", by.y = "ensembl_gene_id")
#remove NA entrez so GSEA_RUN works
entrezdegs <- tidyr::drop_na(entrezdegs, entrezgene_id)
```
GSEA_run with pkd data
```{r}
#sort by highest log fold change to lowest for gsea
entrezdegs <- dplyr::arrange(entrezdegs, desc(log2FoldChange))
#Data is ready for GSEA_RUN. Must include entrezID, log2FoldChange, and the chip index set earlier.
pkd_tf_gsea <- GSEA_run(as.character(entrezdegs$entrezgene_id), entrezdegs$log2FoldChange, chip_index, get.RES = TRUE)
```
Visualization-pkd data
```{r}
#First highlight the significant TFs using TF.highlight. This adds them to the figure legend.
TF.highlight1 <- c("SMARCE1", "E2F4")
#gives the TFs a name for figure legend
names(TF.highlight1) <- c("SMARCE1", "E2F4")
#Sets the viridis colors of the plot. Must be one color code per highlighted TF above. If it doesn't match, will not plot.
col1 <- c("#5ec962", "#440154")
#Scatterplot of the GSEA_RUN results. 
Pre_Cyst_Plot<- plot_ES( pkd_tf_gsea, LFC = entrezdegs$log2FoldChange, specialTF = TF.highlight1, TF_colors = col1, plot_title = "Transcription Factor Enrichment: Pre-Cystic")
#Significant TFs that were highlighted earlier now need labels. This step adds a label at the approximate location of each of the highlighted TFs.
Pre_Cyst_Plot<- Pre_Cyst_Plot %>% add_annotations(x=2800, y=0.46985, xref= "x", yref= "y", xanchor='right', text = "SMARCE1", showarrow = T, arrowhead = 3, arrowsize = .5)
Pre_Cyst_Plot<- Pre_Cyst_Plot %>% add_annotations(x=3943, y=0.42385, xref= "x", yref= "y", xanchor='left', text = "E2F4", showarrow = T, arrowhead = 3, arrowsize = .5)
Pre_Cyst_Plot<- Pre_Cyst_Plot %>% add_annotations(x=4361, y=0.35806, xref= "x", yref= "y", text = "E2F4", xanchor='left', showarrow = T, arrowhead = 3, arrowsize = .5)
#Shows the Plot on Viewer
Pre_Cyst_Plot
```
Visualize plot_RES
```{r}
#Another way to view the results. Function plots a line plot of ES and log2FoldChange
#I chose to plot only the significant TFs. Must include the accession label which can be taken from the scatterplot.
plot_RES(GSEA_result = pkd_tf_gsea, LFC = entrezdegs$log2FoldChange, TF = c("SMARCE1", "E2F4"), Accession = c("ENCSR968QDP.SMARCE1.Hep-G2", "ENCSR000DYY.E2F4.GM12878", "ENCSR000EVL.E2F4.HeLa-S3"))
```
#Check for gene target Overlaps for  drug list
```{r}
#Filter for significant enrichment scores
pkd_tf_gsea_database1_pvalues <- dplyr::filter(pkd_tf_gsea$Enrichment.table, pval.adj < 0.05)
#Remove Unneeded columns
pkd_tf_gsea_database1_pvalues <- pkd_tf_gsea_database1_pvalues[,!(colnames(pkd_tf_gsea_database1_pvalues) %in% c("Accession", "Cell", "Treatment", "Arg.ES", "p.val"))]
#Get rid of repeating TFs
pkd_tf_gsea_database1_pvalues <- dplyr::distinct(pkd_tf_gsea_database1_pvalues, TF, .keep_all = TRUE)
#Create a new object and make it a dataframe
gene_targets_unique_df_TFEA <- as.data.frame(gene_targets_unique)
#Rename columns to match GSEA result column names
gene_targets_unique_df_TFEA <- setNames(gene_targets_unique_df_TFEA, c("TF"))
#merge for matches
target_gene_overlap_TFEA_39 <- merge(pkd_tf_gsea_database1_pvalues, gene_targets_unique_df_TFEA, by = "TF")
View(target_gene_overlap_TFEA_39)
```


```{r}
R.Version()
```
```{r}
installed.packages()
[names(sessionInfo()$otherPkgs), "Version"}]
```

